---
title: "PathwayAnalysisworkshop.Rmd"
author: "rhodgson"
date: "02/06/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages/libraries
```{r}
#BiocManager::install("clusterProfiler")

#BiocManager::install("DOSE")
#BiocManager::install("enrichMAP")


```

Load libraries
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(DESeq2)
library(EnsDb.Mmusculus.v79)
library(pheatmap)
library(RColorBrewer)
library(clusterProfiler)
library(org.Mm.eg.db)
library(DOSE)
library(cowplot)
library(enrichplot)
```

Load files - these are from the DESeq2 workshop on Monday 01/06/20. filter_p05 is filtered on log2FC over 1 and p value <0.05 
allgenes is all genes comparing cd4 v cd8 DKO
```{r}
filter_p05 <- read.csv("/Users/rhodgson/OBDStestdata/Week5/filterbyp05.csv", stringsAsFactors = FALSE)
filter_p01 <- read.csv("/Users/rhodgson/OBDStestdata/Week5/filterbyp01.csv", stringsAsFactors = FALSE)
allgenes  <- read.csv("/Users/rhodgson/OBDStestdata/Week5/allgenes.csv", stringsAsFactors = FALSE)
dim(filter_p01)
dim(filter_p05)
dim(allgenes)
```
First going to look into how to do overrepresentation analysis:
https://yulab-smu.github.io/clusterProfiler-book/
Use function:
enrichGO - need to get ENtrezID

Convert gene ids to entrezid for ech table
As you read in all genes tables need to put in strings as factors
```{r}
edb <- EnsDb.Mmusculus.v79
entrezid_allgenes <- mapIds(edb, keys = allgenes$Ensemble_id, column = "ENTREZID", keytype = "GENEID")
allgenes$entrez_id <- entrezid_allgenes[allgenes$Ensemble_id]
View(allgenes)
```
Do the same for the other 2 tables
```{r}
entrezid_05 <- mapIds(edb, keys = filter_p05$Ensemble_id, column = "ENTREZID", keytype = "GENEID")
filter_p05$entrez_id <- entrezid_05[filter_p05$Ensemble_id]
View(filter_p05)
```

```{r}
entrezid_01 <- mapIds(edb, keys = filter_p01$Ensemble_id, column = "ENTREZID", keytype = "GENEID")
filter_p01$entrez_id <- entrezid_01[filter_p01$Ensemble_id]
View(filter_p01)
```


filter for na genes and pull for entrez id to just get column - can otherwise get table$column in enrichGO
```{r}
genelist_all <- allgenes %>% dplyr::filter(!is.na(entrez_id)) %>% pull(entrez_id)
genelist_p05 <- filter_p05 %>% dplyr::filter(!is.na(entrez_id)) %>% pull(entrez_id)
genelist_p01 <- filter_p01 %>% dplyr::filter(!is.na(entrez_id)) %>% pull(entrez_id)
View(genelist_all)
```

Gene enrichment - "is this set of (DE) genes enriched for each (pathway) geneset, relative to the whole genome"
Give genelist all as the background gene set - this will read as relative to the whole genome
ketype is what we're using as a key
readable = TRUE (sets them to the gene name rather than ens id etc)
```{r}
ego05 <- enrichGO(genelist_p05, OrgDb = org.Mm.eg.db, keyType = "ENTREZID", 
         ont = "ALL",
         universe=as.character(genelist_all),
         pvalueCutoff = 0.05, 
         readable = TRUE)
dim(ego05)
head(ego05, n=50)
```
Same for 0.01

```{r}
ego01 <- enrichGO(genelist_p01, OrgDb = org.Mm.eg.db, keyType = "ENTREZID", 
         ont = "ALL",
         universe=as.character(genelist_all),
         pvalueCutoff = 0.01, 
         readable = TRUE)
dim(ego01)
head(ego01, n=50)
```

Visualisation

```{r}
dotplot(ego05, showCategory=20,  font.size = 7)+ggtitle("dotplot for GO_BP")

```

```{r}
#library(enrichplot)
barplot(ego05, showCategory=20, font.size = 7)+ggtitle("Bar for GO_BP")
```
KEGG enrichment analysis - vectir of genes, function will return the enrichmen KEGG categories with FDR control
```{r}
kegg_05 <- enrichKEGG(gene=genelist_p05,
                      organism='mmu',
                      pvalueCutoff = 0.05)
barplot(kegg_05, showCategory=20, font.size = 7)
```


```{r}
kegg_01 <- enrichKEGG(gene=genelist_p01,
                      organism='mmu',
                      pvalueCutoff = 0.01)
barplot(kegg_01, showCategory=20, font.size = 7)
```

```{r}
p1 <- emapplot(ego05, font.size=6, layout= "nicely", showCategory = 15, line_scale=5)
p1
p2 <- emapplot(ego01, font.size=6, layout= "nicely", showCategory = 10)
cowplot::plot_grid(p1, p2, ncol=2, labels=LETTERS[1:2])
```

```{r}
upsetplot(ego01)
```


```{r}

p3 <- barplot(kegg_05, showCategory=8, font.size = 4)+ ggtitle("KEGG0.05")
p4 <- dotplot(kegg_01, showCategory=30, font.size = 4) + ggtitle("KEGG0.01")
cowplot::plot_grid(p3, p4, ncol=2, labels=LETTERS[1:2])
```

Compare gene clusters functional profile:
Given a list of gene set, this function will compute profiles of each gene cluster.
gene_list_low005 gave me the genes most downregulated over -1 in previous script


```{r}
gene_list_low005 <- read.csv("/Users/rhodgson/OBDStestdata/Week5/gene_list_low005.csv", stringsAsFactors = FALSE)

gene_list_low005 <- gene_list_low005 %>%  dplyr::pull(x) 
class(gene_list_low005)
View(gene_list_low005)
```

```{r}
#Compare cluster bit:

xx <- compareCluster(geneClusters = list(up=as.character(genelist_p05), down= as.character(gene_list_low005)), fun="enrichKEGG", "org.Mm.eg.db",organism="mmu", pvalueCutoff=0.05, universe=as.character(genelist_all))
dotplot(xx, showCategory=5)

```
Do the same for 0.01

```{r}
gene_list_low001 <- read.csv("/Users/rhodgson/OBDStestdata/Week5/gene_list_low001.csv", stringsAsFactors = FALSE)

gene_list_low001 <- gene_list_low001 %>%  dplyr::pull(x) 
class(gene_list_low001)
x2 <- compareCluster(geneClusters = list(up=as.character(genelist_p01), down= as.character(gene_list_low001)), fun="enrichKEGG", "org.Mm.eg.db",organism="mmu", pvalueCutoff=0.01, universe=as.character(genelist_all))
dotplot(x2, showCategory=5)
```
Tells you pathways enriched in different clusters - would take top upregulated genes from clusters
Would compare 

```{r}
x1 <- compareCluster(geneClusters = list(up=as.character(genelist_p05), down=as.character(gene_list_low005)),OrgDb="org.Mm.eg.db", fun="enrichGO", pvalueCutoff=0.05, universe = as.character(genelist_all))

x1 <- dotplot(x1, showCategory=5)
x1
```


Validation:
```{r}
cc <- compareCluster(genelist_p01, fun="enrichKEGG",
                     organism="mmu", pvalueCutoff=0.01)
#p1 <- emapplot(cc)
#p2 <- emapplot(cc,legend_n=2) 
##p3 <- emapplot(cc,pie="count")
#p4 <- emapplot(cc,pie="count", pie_scale=1.5, layout="kk")
#cowplot::plot_grid(p1, p2, p3, p4, ncol=2, labels=LETTERS[1:4])
```


#GO gene set enrichment

```{r}
#gse_05 <- gseGO(geneList=genelist_p05,
              OrgDb        = org.Mm.eg.db,
              ont          = "CC",
              nPerm        = 1000,
              minGSSize    = 100,
              maxGSSize    = 500,
              pvalueCutoff = 0.05,
              verbose      = FALSE)
```


#KEGG Gene Set Enrichment Analysis
input is gene list (3.1 of manua tells you what it is)
gene list is numeric vector of fold cnages - table of all genes from analysis (ie. allgenes)
Trying to sort out the inpt file here - making sure 
```{r}
length(allgenes$Ensemble_id)
length(which(is.na(allgenes$entrez_id)))
library(dplyr)
allgenes_f <- allgenes %>% dplyr::filter(!is.na(entrez_id)) 
head(allgenes_f)

```
```{r}
#%>% pull(entrez_id)
results_all_FC <- allgenes_f$log2FoldChange

names(results_all_FC) <- allgenes_f$entrez_id
head(results_all_FC)
table(is.na(names(results_all_FC)))


#sort by fold change
results_all_FC <- sort(results_all_FC, decreasing = TRUE)
head(results_all_FC)

results_all_FCdup <- results_all_FC %>% dplyr::filter(!(duplicated(results_all_FC))) 

```

Check duplicate entrezids
```{r}
duplicated <-table(duplicated(results_all_FC))
head(duplicated)


```

```{r}
kk2 <- gseKEGG(geneList = results_all_FC,
              organism = "mmu")
head(kk2)

#xx <- gseKEGG(geneList = all_scores, organism = "mmu")
#as.data.frame(xx)
```

```{r}
ego3 <- gseGO(geneList     = results_all_FC,
              OrgDb        = org.Mm.eg.db,
              ont          = "ALL",
              nPerm        = 1000,
              minGSSize    = 100,
              maxGSSize    = 500,
              pvalueCutoff = 0.05,
              verbose      = FALSE)

as.data.frame(ego3)

p8 <- gseaplot(ego3)
p8
```

