---
title: "StatsinR.Rmd"
author: "rhodgson"
date: "28/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Exercise 1:
Generate a vector of 1000 normally distributed values with mean 10 and standard deviation 5.
Print summary statistics about those values.
```{r}
x <- rnorm(n=1000, mean=10, sd=5)
summary(x)
```
Verify the mean and standard deviation. Inspect the deciles of those values. - quantile gives you the first 0-25-50-75-100 quantiles
Orders your data by section
```{r}
mean(x)
sd(x)
quantile(x)
#deciles 

#quantiles(x, probs= seq(0,1,0.1)
```
Visualise the distribution of those values. Draw vertical lines to indicate the mean and 1 standard deviation either side. Bonus point if the lines are colored.
1. Using base R.
```{r}
hist(x, breaks=30)
abline(v=mean(x), col="blue")
abline(v=(mean(x)+sd(x)), col="red")
abline(v=(mean(x)-sd(x)), col="red")
```


2. Using ggplot .
```{r}
library(ggplot2)
```

```{r}
x_df <- as.data.frame(x)
ggplot(x_df, aes(x=x))+
  geom_histogram()+
  geom_vline(aes(xintercept=mean(x)),color="blue", linetype="dashed", size=1)+
  geom_vline(aes(xintercept=sd(x)),color="red", linetype="dashed", size=1)+
  geom_vline(aes(xintercept=mean(x)+sd(x)),color="red", linetype="dashed", size=1)

```

Verify that approximately 64% and 95% of the values are within 1 and 2 standard deviations of the mean, respectively.

```{r}
percent1sd <- sum(x_df$x>=mean(x)-sd(x) & x_df$x<= mean(x)+sd(x))/(length(x_df$x))*100 
    
percent1sd


percent_2sd <- sum(x_df$x>=mean(x)-2*sd(x) & x_df$x<=mean(x)+2*sd(x))/(length(x_df$x))*100
percent_2sd
```

Generate a new vector with a lot more values (e.g., one million). Draw again a histogram. Does the distribution look better? worse?
```{r}
x_large <- rnorm(n = 1000000, mean = 10, sd = 5)
x_large_data <- data.frame(x=x_large)
ggplot(x_large_data, aes(x=x_large))+
  geom_histogram(bins = 100)+
  geom_vline(xintercept=mean(x), color="blue")+
  geom_vline(xintercept=mean(x) + c(-1, 1)*sd(x), color="red")

```


#Statistical testing for iris data set
In the iris dataset, visualise the distribution of sepal length stratified by species.

```{r}
data(iris)
ggplot(iris, aes(x=Sepal.Length))+geom_histogram()+
facet_wrap(~ Species)

```
Overlay by specifying column is 1
```{r}
data(iris)
ggplot(iris, aes(x=Sepal.Length))+
  geom_histogram()+
  facet_wrap(~ Species, ncol=1)
```


Print summary statistics for each column in the dataset.
How many species are there in the dataset? What are their names? How many observations do
we have for each species?
```{r}
summary(iris)
```

Is the sepal length normally distributed overall? Within each species?
Can I use shapiro test here

```{r}
library(tidyverse)
Sepallength <- iris %>% group_by(Species) %>% select() 

shapiro.test(iris$Sepal.Length) 
```
for filter(species == "A"):
```{r}
#species <- levels(iris$Species)
#species
iris_setosa <- iris %>% filter(Species == "setosa")
iris_setosa <- iris_setosa$Sepal.Length
shapiro.test(iris_setosa)

#virginica
iris_virginica <- iris %>% filter(Species == "virginica")
iris_virginica <- iris_virginica$Sepal.Length
shapiro.test(iris_virginica)
#Do this for versicolor
iris_versicolor <- iris %>% filter(Species == "versicolor")
iris_versicolor <- iris_versicolor$Sepal.Length
shapiro.test(iris_versicolor)
```
```{r}
iris_virginica
```

Fancy for loop doing the same thing
ie. for each species of the levels Species, print the species, filter by Species, pull the Sepal.Length column and perform the shapiro test on each set
```{r}
for (species in levels(iris$Species)) {
  print(species)
  iris %>%
    filter(Species == species) %>%
    pull(Sepal.Length) %>%
    shapiro.test() %>%
    print()
}
```

```{r}
shapiro.test(iris$Sepal.Length)
```
Looks fine and normally distributed

Next:
Is there a significant variation between the sepal length between the different species? 

Use Kevin's for loop to do this (feeding into the var() function)
No this isnt right, we want to use anova here instead
```{r}
for (species in levels(iris$Species)) {
  print(species)
  iris %>%
    filter(Species == species) %>%
    pull(Sepal.Length) %>%
    var() %>%
    print()
}
```
Compute the analysis of variance res.aov <- aov(weight ~ group, data = my_data) # Summary of the analysis summary(res.aov)

stars show yes species affected distribution of sepal length 
quick way of showing distribution 
```{r}
anova <- aov(Sepal.Length ~ Species, data = iris)
anova
summary(anova)
```
Easy way to compare separate easily (previously made vectors of just sepal_length for each one)
```{r}
t.test(x= iris_setosa, y=iris_versicolor)
```

Do setosa and versicolor species have significantly different sepal length?
can use the same syntax of anova test (the ~ is a formal says left is predictor and right side is the independent)
filter N then test
```{r}
iris_sv <- iris %>% filter(Species == "setosa"|Species == "virginica") 

t.test(Sepal.Length ~ Species, data=iris_sv)
```

Pulling only works on one column so we lost the species info here - doing a t test on one sample. This doesn't work 
```{r}
iris %>% filter(Species == "setosa" | Species == "virginica") %>% pull(Sepal.Length) %>%
  t.test()
```
. means data coming from the pipe. 
```{r}
iris %>% filter(Species == "virginica"|Species == "setosa") %>% t.test(Sepal.Length ~ Species, data= . )

```
#Multiple testing using ALLL Microarray data set
Expression set object - container for complex genomic experimnet - all the info - split out the expression variable into one object and metadata into another. Should allow you to subset easier (not sure if it does)

I

Use the ALL microarray gene expression dataset in the ALL package. The normalized expression data is stored in exprs(ALL) . Sample metadata is stored in pData(ALL)
```{r}
BiocManager::install("ALL")
library(ALL)
```


```{r}
data(ALL)
head(exprs(ALL))
```


Use the following code to select samples from B-cell lymphomas harboring the BCR/ABL translocation and from lymphomas with no observed cytogenetic abnormalities (NEG).

```{r}
bcell = grep("^B", as.character(ALL$BT))
moltyp = which(as.character(ALL$mol.biol) %in% c("NEG", "BCR/ABL"))
ALL_bcrneg = ALL[, intersect(bcell, moltyp)]
ALL_bcrneg$mol.biol = factor(ALL_bcrneg$mol.biol)
str(ALL_bcrneg)
```

Assume all normally distributed
Test each microarray probeset between patients who achieved remission and those who were refractory to treatment.

Test each row in dataframe, see if its signatificantly differ
Take first row i matrix and test between two groups

1 probe (row)
2 groups of samples, 
```{r}
exprs(ALL_bcrneg[1])

#View(a)
```
Filter first row (1 gene) and give a new column on mol_status is the molbiol col
```{r}
ALL_df <- data.frame(gene =exprs(ALL_bcrneg)[1,], mol_status = ALL_bcrneg$mol.biol)

ALL_df
```
Now to do t test
```{r}
t<- t.test(gene ~ mol_status, data= ALL_df)
str(t)
```
Class htest - allows oyu to give summary
Can also look in this and extract certain parts
```{r}
pval_1 <- t$p.value
pval_1
```
Now want to do this for every gene in data set using sapply function

Define our test_probe function so we will give a vector of values in the row
x will become our values
```{r}
test_probe <- function(x){
  ALL_df <- data.frame(gene =x, mol_status = ALL_bcrneg$mol.biol) 
  t <- t.test(gene ~ mol_status, data= ALL_df)
  p_x <- t$p.value
  return(p_x)
}
```


Apply function

```{r}

pvalues_combined <- apply(exprs(ALL_bcrneg), 1, test_probe)
head(pvalues_combined)
```



```{r}
#hist(pvalues_combined)
pvalues_combined_df <- as.data.frame(pvalues_combined)
colnames(pvalues_combined_df)
ggplot(pvalues_combined_df, aes(x=pvalues_combined))+geom_histogram()
```
For each p values, is it less than 0.05 and how many are there?
This is a good sign there there is something to look at 
```{r}
length(pvalues_combined[pvalues_combined<0.05])
```
Dharam's way of doing this
```{r}
pvalues_combined_df %>%filter(pvalues_combined <= 0.05) %>% nrow()
```

Correct p-values for multiple testing. How many probesets remain significant? Using the Benjiman Hochberg

```{r}

p_adjust <- p.adjust(pvalues_combined, method = "BH", n = length(pvalues_combined))
plot(pvalues_combined, p_adjust)
```
How many probset are still significant?
```{r}
length(p_adjust[p_adjust<0.05])
hist(p_adjust)
abline(v=0.05, col="red")
```
Plot the expression of the most signficant probeset in the two groups of samples.

Find the value with smallest p number
```{r}
p_adjust[which.min(p_adjust)]
```
Mapping with gene entries etc
maps probes of microarray to gene identifiers
```{r}
#BiocManager::install("hgu95av2.db")
library(hgu95av2.db)
```
Now find out what that gene is called
Bonus point: Does the most signi cant probeset map to a gene? If so, which one?
```{r}
AnnotationDbi::select(hgu95av2.db, "1636_g_at", columns = "SYMBOL")

```



Visualise the distribution of unadjusted p-values for the two probesets with high and low variance.
Grab that particular probeset
```{r}
ABLprobe <- data.frame(gene =exprs(ALL_bcrneg)["1636_g_at",], mol_status1 = ALL_bcrneg$mol.biol)
View(ABLprobe)
```

Do the groups have different expression of this gene?
ggplot
```{r}
ggplot(ABLprobe, aes(x=mol_status1, y=gene))+geom_boxplot(col="red")+
  geom_jitter()

```
How would you use variance be used to reduce the burden of multiple testing correction
Calculte variance for each row of a dataframe
```{r}
library(matrixStats)
```

Calculates vairance of each row
Which rowvars are less than the median variance 

I dont particular understand ths
Its a way of looking at variance of the individual probes of a dataset to only pick the ones that have higher variance 

```{r}
rowvars <- rowVars(exprs(ALL_bcrneg)[names(pvalues_combined),]) #gives row variance in correct order
med <- median(rowvars)
row_vars_low <- pvalues_combined[rowvars<med]
hist(row_vars_low, breaks =100)


```

```{r}
row_vars_high_p <- pvalues_combined[rowvars>med]
hist(row_vars_high_p, 100)
```

